% Survey Instrument
%Everything should be available at: ZENODO.

The survey instrument to gather the students' feedback about the course is presented below. 

\subsection{Presentation}
\textit{Dear Student,}

\textit{This survey aims to evaluate the course MATE28-Topics in Software Engineering IV, which dealt with Practices and Reflections on Contribution in Free/Libre Open Source Software Projects (FLOSS) in 2021.2.}

\textit{Completing the questionnaire may take 10 to 15 minutes. Your responses will be treated completely anonymously.}
 
\textit{Thanks for staying with us until the end of the course. Now, your participation is very important in this evaluation and self-assessment process!}

\textit{The SEED.br Group Team.}

\subsection{Consent Form} 
(Ommitted)

\subsection{Participant Profile}

\begin{enumerate}
    \item Gender
    \item Age
    \item Education level
    \item Professional experience
    \item General comments about experience (open-ended question).
\end{enumerate}

\subsection{Isolation Context}
(COVID-19. Ommitted)

\subsection{Self-evaluation}

\begin{enumerate}
    \item Selected OSS project: name and GitHub URL (open-ended)
    \item I can grasp the complexity and difficulties in building models of a software project in public repositories (Likert scale)
    \item I can identify the content discussed in the classroom in a real-world, medium-sized software. (Likert scale)
    \item I can perform activities similar to those performed during the course in another software project. (Likert scale)
    \item The experience contributed to the improvement of my professional performance (current or future). (Likert scale)
    \item The experience contributed to the development of the proactivity skill. (Likert scale)
    \item The experience contributed to the development of problem solving skills. (Likert scale)
    \item The experience contributed to the development of communication skills. (Likert scale)
    \item The experience helped me to have contact with code developed by third parties. (Likert scale)
\end{enumerate}

\subsection{Experience Report}

Open-ended questions.

\begin{enumerate}
\item How was your experience in the OSS project selection process (positive and negative points)?

\item How was your experience in carrying out the characterization of the selected project (positive and negative points)?

\item How was your experience in carrying out the practical activity on bug resolution (positive and negative points)?

\item Describe your experience in building the ``journal'' (positive and negative points) 

\item Describe your experience with the OSS community (commits, community feedback, etc.)

\item Did the course meet your expectations? Please comment.

\item Were the strategies used in the classes during the semester adequate? Please comment on positive points and points for improvement. Suggestions are welcome.

\item General comments (at will).

\end{enumerate}
