The template for the initial characterization report on the selected OSS project.

\subsection{Part I - Project description}
\label{part-i-description-of-the-project}

\begin{enumerate}
\def\labelumi{\arabic{enumi}.}
\item
   Name
\item
   Web site
\item
   Description
\end{enumerate}

\subsection{Parte II - Initial characterization}
\label{parte-ii---reconhecimento-inicial}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  How long has the project existed? Briefly describe the history of the project.
\item
  What is the project license?
\item
  What is the project size (lines of code, classes)?
\item
  Has the project size grown in recent versions?
\item
  Is there recent activity on the project?
\item
  Is there collaboration from companies on the project?
\item
  What are the technologies used?
\end{enumerate}

\subsection{Parte III - Issues characterization}
\label{parte-iii---identificauxe7uxe3o-de-tarefas}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{10}
\item
  Where is the source code repository? Which version control tool is used?
\item
  What kind of bug repository is used and where is it located?
\item
  Is there documentation to help new contributors?
\item
  Are there any bugs that are marked as easy to fix?
\item
  Does the source code contain issues lists?
\item
  Are there any comments containing TODO, FIXME, etc in the source code? What kind of need are placed in these comments?
\end{enumerate}

\subsection{Parte IV - Technical characterization}\label{parte-iv---caracterizauxe7uxe3o-tuxe9cnica}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{17}
\item
   How many developers participated in the project in the last 6 months? And from the beginning?
\item
   What is the version release policy?
\item
   Does the project have an automated test suite?
\item
   Are there parts of the source code that are problematic? Justify.
\item
   Does the code contain comments? Are there parts that need further comment?
\item
   Can design patterns and architectural patterns be identified from the code? Which ones did you identify?
\item
   How was the project source code obtained?
\item
   What did you need to install on your system to make the project
   compile/run? Was the documentation provided by the project sufficient?
\end{enumerate}
